### 2020-09-15 Manuscript/manuscript.tex
smalls edits to tectonic into and additions to glossary

Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic following its ca. 1.8 Ga amalgamation. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic-Mesoproterozoic Nuna and Neoproterozoic Rodinia supercontinents. There is a rich record of Precambrian paleomagnetic poles from Laurentia, as well as an extensive and well-documented geologic history of tectonism. These geologic and paleomagnetic records are increasingly better constrained geochronologically and are both key to evaluating and developing paleogeographic models. These data from Laurentia provide strong-support for mobile lid plate tectonic processes operating continuously over the past 2.2 billion years.

@@

Laurentia refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic ca. 1.8 Ga, as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep[e.g.]{Skipton2016a, Weller2017a}. A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca. 1.89 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

As for the suturing of the Wyoming province to Laurentia, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to Hearne and Superior while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming and Medicine Hat provinces were not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016b}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a, Hrncir2017a}, as well as high-grade metamorphism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between Wyoming and Superior. Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast. Overall, the collision of these Archean microcontinents between ca. 1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton (Fig. \ref{fig:Laurentia_map}).

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a} with additions for Greenland based on \citealp{St-Onge2009a}). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel.}}}
\label{fig:Laurentia_map}
\end{figure}

@@

The collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna in the paleogeographic model framework of \cite{Pehrsson2015a}. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}; Fig. \ref{fig:Superior_Slave_recons}). The paleogeographic mode of \cite{Pehrsson2015a} posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo, S\~ao Francisco, West Africa, and Amazonia. In this volume, \cite{Elming2021a} put forward an alternate scenario wherein Laurentia, Baltica and Siberia become conjoined at the time of Laurentia amalgamation forming the core of Nuna (as in \citealp{Evans2011a}). In the \cite{Elming2021a} model, this core subsequently grows to be a semi-supercontinent with India and Australia, however Amazonia, West Africa, Congo and S\~ao Francisco cratons remain independent from Nuna.

@@

\bibliographystyle{gsabull}
\footnotesize\bibliography{../../references/allrefs}

@@

\noindent\textbf{Archean province } A contiguous area of Archean continental lithosphere typically surrounded by Proterozoic orogens inferred to be suture zones (e.g. Superior province). \\

@@

@@

\noindent\textbf{Laurentia } The cratonic core of the North America continent that formed through the amalgamation of Archean provinces in the Paleoproterozoic.

@@

@@

@@

@@

@@

@@

@@

@@

@@

\noindent\textbf{Thelon orogen } An orogen resulting from a Paleoproterozoic collision orogeny between the Slave and Rae provinces.

\noindent\textbf{Wopmay orogen } An orogen resulting from collision between the Hottah terrane, a continental magmatic arc, and the west margin of the Slave province.

>>>>

Laurentia is the craton that forms the Precambrian core of North America and was a major continent throughout the majority of the Proterozoic following its ca. 1.8 Ga amalgamation. The paleogeographic position of Laurentia is key to the development of reconstructions of Proterozoic paleogeography including the Paleoproterozoic to Mesoproterozoic supercontinent Nuna and latest Mesoproterozoic to Neoproterozoic supercontinent Rodinia. There is a rich record of Precambrian paleomagnetic poles from Laurentia, as well as an extensive and well-documented geologic history of tectonism. These geologic and paleomagnetic records are increasingly better constrained geochronologically and are both key to evaluating and developing paleogeographic models. These data from Laurentia provide strong-support for mobile lid plate tectonic processes operating continuously over the past 2.2 billion years.

@@

Laurentia refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca. 1.8 Ga), as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989a, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989a}; the Snowbird orogen associated with ca. 1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the southern Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

As for the suturing of the Wyoming province to Laurentia, many models posit that it was conjoined with Hearne and associated provinces at the time of the Trans-Hudson orogeny \citep[e.g.][]{St-Onge2009a, Pehrsson2015a} or was proximal to the Hearne and Superior provinces while still undergoing continued translation up to ca. 1.80 Ga \citep{Whitmeyer2007a}. A contrasting view has been been proposed that the Wyoming and Medicine Hat provinces were not conjoined with the other Laurentia provinces until ca. 1.72 Ga \citep{Kilian2016b}. This interpretation is argued to be consistent with geochronological constraints on monazite and metamorphic zircon indicating active collisional orogenesis associated with the Big Sky orogen on the northern margin of the craton as late as ca. 1.75 to 1.72 Ga \citep{Condit2015a} and ca. 1.72 tectonomagmatic activity in the Black Hills region \citep{Redden1990a}. However, the evidence for earlier orogenesis ca. 1.78 to 1.75 in the Black Hills \citep{Dahl1999a, Hrncir2017a}, as well as high-grade metamorphism as early as ca. 1.81 Ga in the Big Sky orogen \citep{Condit2015a}, may support the interpretation of \citet{Hrncir2017a} that ca. 1.72 Ga activity is a minor overprint on ca. 1.75 terminal suturing between the Wyoming and Superior provinces. Regardless, in both of these interpretations, Wyoming is a later addition to Laurentia with final suturing post-dating ca. 1.82 Ga amalgamation of Archean provinces with the Trans-Hudson orogen further to the northeast. Overall, the collision of these Archean microcontinents between ca. 1.9 and 1.8 Ga led to rapid amalgamation of the majority of the Laurentia craton (Fig. \ref{fig:Laurentia_map}).

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Fig1_map.pdf}
\caption{\small{\textbf{Simplified map of Laurentia showing the location of Archean provinces (labeled with text) and younger Paleoproterozoic and Mesoproterozoic crust (simplified from \citealp{Whitmeyer2007a} with additions for Greenland based on \citealp{St-Onge2009a}). Proterozoic orogens are labeled with \textit{italicized text} (or. ? orogen). The localities from which the compiled Precambrian paleomagnetic poles were developed are shown and colored by age. The circles (A rated poles) and squares (B rated poles) have been assessed by the Nordic workshop panel.}}}
\label{fig:Laurentia_map}
\end{figure}

@@

The collisions of provinces and terranes leading up to the Trans-Hudson orogeny mark the initial phase of assembly of the supercontinent Nuna in the paleogeographic model framework of \cite{Pehrsson2015a}. The Trans-Hudson orogeny itself is taken to be the terminal collision associated with the closure of the Manikewan Ocean that had previously been a large oceanic tract separating the Superior province from the composite Slave+Rae+Hearne+North Atlantic provinces (often referred to as the Churchill domain or plate; e.g. \citealp{Skipton2016a, Weller2017a}; Fig. \ref{fig:Superior_Slave_recons}). The paleogeographic mode of \cite{Pehrsson2015a} posits that this period of terminal collision not only resulted in the amalgamation of Laurentia, but was also associated with the assembly of the supercontinent Nuna that is hypothesized to include other major Paleoproterozoic cratons including Baltica, Siberia, Congo, S\~ao Francisco, West Africa, and Amazonia. In this volume, \cite{Elming2021a} put forward an alternate scenario for Nuna paleogeography. In their model,  Laurentia, Baltica and Siberia become conjoined at the time of Laurentia amalgamation forming the core of Nuna (as in \citealp{Evans2011a}). This core then subsequently grows to be a semi-supercontinent with India and Australia, however Amazonia, West Africa, Congo and S\~ao Francisco cratons remain independent from Nuna.

@@

\singlespacing
\bibliographystyle{gsabull}
\footnotesize\bibliography{../../references/allrefs}

@@

\noindent\textbf{Archean province } A contiguous area of Archean continental lithosphere typically surrounded by Proterozoic orogens inferred to be suture zones (e.g. Superior province).

\noindent\textbf{Canadian shield } Large area of exposed Precambrian rock of Laurentia, or rock covered by thin soil, that is well-exposed due to Pleistocene glaciation erosion

@@

\noindent\textbf{craton } The stable and relatively immobile continental lithosphere in the interior of continents. In this chapter, craton is predominantly used in reference to Laurentia which formed through the collision of Archean provinces and grew further through subsequent accretionary and collisional orogenesis.

@@

\noindent\textbf{Grenvillian orogen } An orogen resulting from collision orogeny between Laurentia and conjugate continent(s) near the end of the Mesoproterozoic.

\noindent\textbf{Hearne province } An Archean province of Laurentia extending from southern Alberta, Canada to Hudson Bay. It is framed by the Rae province to the northwest and the Trans-Hudson orogen to the southeast. It is also referred to as the Hearne craton.

\noindent\textbf{Laurentia } The Precambrian cratonic core of the North America continent and Greenland that formed through the amalgamation of Archean provinces in the Paleoproterozoic.

@@

\noindent\textbf{Medicine Hat province } An Archean province of Laurentia extending from northern Montana, USA into southern Alberta and Saskatchewan, Canada. It is framed by a suture with the Hearne province to the north, the Trans-Hudson orogen to the east, and the Great Falls tectonic zone to the south. It is also referred to as the Medicine Hat block

@@

\noindent\textbf{Meta Incognita province } A province of Archean basement rocks that comprises most of southern Baffin Island. It is also referred to as the Meta Incognita microcontinent.

@@

\noindent\textbf{Nagssugtoqidian orogen } An orogen resulting from the Paleoproterozoic collision between the Rae and North Atlantic provinces.

@@

\noindent\textbf{North Atlantic province } An Archean province of Laurentia in southernmost Greenland and northeastern Labrador, Canada. It is also referred to as the North Atlantic craton.

@@

\noindent\textbf{Penokean orogen } An orogen resulting from Paleoproterozoic accretion of an oceanic arc and the Marshfield terrane continental block along the southern margin of the Superior province.

@@

\noindent\textbf{Proterozoic } A geologic eon spanning from 2,500 to 541 million years ago.

@@

\noindent\textbf{Rae province } An Archean province of Laurentia extending the region of Lake Athabasca northeast to northern Baffin Island in arctic Canada. It is framed by the Thelon orogen to the west, the Taltson orogen to the southwest and the Hearne province to the east. It is also referred to as the Rae craton.

@@

\noindent\textbf{Slave province } An Archean province of Laurentia extending to the north from the region of Great Slave Lake in northern Canada. It is framed by the Thelon orogen to the east and the Great Bear Arc to the west. It is also referred to as the Slave craton.

\noindent\textbf{Snowbird orogen } An orogen resulting from Paleoproterozoic collision between the Rae and Hearne provinces prior to the Trans-Hudson orogeny. The Snowbird tectonic zone is part of the orogen.

@@

\noindent\textbf{Thelon orogen } An orogen resulting from the Paleoproterozoic collision orogeny between the Slave and Rae provinces.

\noindent\textbf{Torngat orogen } An orogen resulting from the Paleoproterozoic collision orogeny between the Meta Incognita and North Atlantic provinces.

\noindent\textbf{Trans-Hudson orogen } An orogen resulting from the Paleoproterozoic collision between the composite Slave+Rae+Hearne provinces and the Superior province.

\noindent\textbf{Wopmay orogen } An orogen resulting from Paleoproterozoic collision between the Hottah terrane, a continental magmatic arc, and the west margin of the Slave province.

\noindent\textbf{Wyoming province } An Archean province of Laurentia underlying much of Wyoming, USA and southeast Montana, USA. It is framed by the Trans-Hudson orogen to the east (sometimes referred to the Black Hills orogen within the USA) and the Great Falls tectonic zone to the north. It is also referred to as the Wyoming craton.
